---
title: "Getting Started with glystats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with glystats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

When working with omics data, it is often useful to begin with a small set of
reliable analyses: differential expression, PCA, survival analysis, and more.
`glystats` brings these methods together through one consistent,
user-friendly interface.

**A useful feature:** Every `glystats` analysis accepts
`glyexp::GlycomicSE()` and `glyexp::GlycoproteomicSE()` containers,
the unified data interfaces used across the `glycoverse` ecosystem.

If these containers are new to you, the
[glyexp introduction](https://glycoverse.github.io/glyexp/articles/glyexp.html)
provides a useful overview.

```{r setup, message = FALSE}
library(glystats)
library(glyexp)
library(glyread)
library(glyclean)
library(dplyr)
```

## Explore the data

We will start by exploring the demo dataset:

```{r message = FALSE}
# Here we use `glyread` to read in the data and use `glyclean` to perform preprocessing
exp <- read_pglyco3_pglycoquant("glycopeptides.list", sample_info = "sample_info.csv") |>
  auto_clean()
```

```{r}
exp
```

The result is a `glyexp::GlycoproteomicSE()` containing 12 samples and 263
glycoforms, which is enough to illustrate the main workflow.

```{r}
tibble::as_tibble(rowData(exp), rownames = "variable")
```

The variable information tibble contains descriptive information for each
glycoform, including the protein, glycosylation site, and glycan structures.

```{r}
tibble::as_tibble(colData(exp), rownames = "sample")
```

The sample information tibble includes a `group` column. In this dataset,
"H" represents healthy samples, "M" hepatitis, "Y" cirrhosis, and "C"
hepatocellular carcinoma. We will use this column for the comparisons below.

## Use a consistent interface

`glystats` uses a consistent naming pattern: `gly_` followed by the analysis
name. For example, `gly_ttest()` runs a t-test and `gly_pca()` performs PCA.

RStudio's auto-completion can help you discover the available functions.

We can use ANOVA to identify glycoforms that differ between groups:

```{r}
anova_res <- gly_anova(exp)
```

`gly_anova()` follows the `glycoverse` column conventions, detects the `group`
column in the sample information, and fits an ANOVA model for each glycoform.

## Work with the results

All `glystats` functions return a consistent list with two main components:

- `tidy_result`: Analysis-ready tibbles in tidy format
- `raw_result`: The original result objects from the underlying statistical functions

For `gly_anova()`, the `tidy_result` contains two informative tibbles:
`main_test` and `post_hoc_test`.

You can use `get_tidy_result()` to get the tidy result tibble:

```{r}
get_tidy_result(anova_res, "main_test")
```

`gly_anova()` also adds the descriptive columns from the variable tibble to the
tidy results. Use the `add_info` parameter when you want to control this
behavior.

The `raw_result` houses two lists of models - one for the main test,
one for post hoc comparisons:

```{r}
names(get_raw_result(anova_res))
```

`get_tidy_result()` and `get_raw_result()` can also be used in a pipe:

```{r}
exp |>
  gly_anova() |>
  get_tidy_result("main_test") |>
  filter(p_adj < 0.05)
```

## Available analyses

The package includes the following analyses for glycomics and glycoproteomics
data:

- **Differential expression analysis:**
  - `gly_ttest()`: Two-sample t-test
  - `gly_wilcox()`: Wilcoxon rank sum test
  - `gly_anova()`: One-way ANOVA
  - `gly_kruskal()`: Kruskal-Wallis rank sum test
  - `gly_limma()`: Linear models for microarray data (limma)
  - `gly_ancova()`: ANCOVA (Analysis of Covariance)
  - `gly_linear_model()`: Formula-based moderated linear models
  - `gly_set_test()`: correlated variable-set testing
  - `gly_fold_change()`: Calculate fold change

- **Dimensionality reduction:**
  - `gly_pca()`: Principal component analysis
  - `gly_tsne()`: t-distributed stochastic neighbor embedding (t-SNE)
  - `gly_umap()`: Uniform manifold approximation and projection (UMAP)
  - `gly_oplsda()`: Orthogonal partial least squares discriminant analysis (OPLS-DA)
  - `gly_plsda()`: Partial least squares discriminant analysis (PLS-DA)

- **Clustering:**
  - `gly_kmeans()`: K-means clustering
  - `gly_hclust()`: Hierarchical clustering

- **Survival analysis:**
  - `gly_cox()`: Cox proportional hazards model

- **Additional tools:**
  - `gly_cor()`: Correlation analysis
  - `gly_roc()`: Receiver operating characteristic (ROC) analysis
